Supplemental Materials: \emph{Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours}

### 2015-03-18 Publications/2015-NeuroImage/supplemental.tex
one more iteration to second round
automatically generate supplemental figures of real data
update supplementals

% @Author: Oscar Esteban
% @Date:   2015-02-20 15:40:09
% @Last Modified by:   Oscar Esteban
% @Last Modified time: 2015-02-27 15:21:03

@@

% *** MISC UTILITY PACKAGES ***
%\usepackage{fancyhdr}
\usepackage{multicol}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs} % to use \toprule
\usepackage{makeidx}
%\usepackage[scale=2.0]{ccicons}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[marginparwidth=1.2in]{geometry}
\usepackage{changepage}
\usepackage{epigraph}
\usepackage[toc,nomain,acronym,shortcuts,translate=false]{glossaries}
\usepackage[hyphens]{url}
\usepackage[breaklinks,hidelinks]{hyperref}
\usepackage[hyphenbreaks]{breakurl} %fixes boxes spanning through pages

\renewcommand\thesection{S\arabic{section}}
\renewcommand\thesubsection{S\arabic{section}.\arabic{subsection}}

@@

@@

\maketitle

@@

\paragraph{Efficient interpolation using sparse matrices}
During the registration process, every iteration requires computing the product of all the speeds
  $\mathbf{s}'_i$ computed at the current position $\mathbf{v}'_i$ by the corresponding weights
  $\psi_k(\mathbf{v}_i)$ of interpolating functions (Eq. (15)).
As these weights can be computed once in the beginning of the process and they do not change along
  it, and they are different from zero only in the proximity of the corresponding control node $k$,
  they can be pre-cached in a sparse matrix.

@@

\subsection{Convergence evidencing}\label{sec:convergence_evidence}
In order to track the evolution of the registration process, several internal variables
  are saved in the JSON log-file.
Using the JSON log-file as input for the \emph{nipype Interface}
  \codeinline{ACWEReport}, it is straightforward to obtain
  a visual assessment document presenting the convergence.
Online checking is also possible as the algorithm writes to the standard output as well.

\begin{pythoncode}
from pyacwereg.interfaces.acwereg import ACWEReport
report = ACWEReport()
report.inputs.in_log = `myprefix.log'
ifresult = report.run()
\end{pythoncode}

\begin{figure*}[!ht]
	\includegraphics[width=\textwidth]{figures/Suppl-figure02.pdf}
	\caption{The evolution of the registration and segmentation process can be
	  checked using the \emph{Convergence report},
	  easily generated using the appropriate \emph{nipype Interface}.
	The report comprehends several plots tracking the evolution of the algorithm and several
	  features to help researchers tune up the algorithm in their application.}%
	  \label{fig:convreport}
\end{figure*}

@@

\section{Extended results on phantom data}

\section{Extended results on real data}

>>>>

% @Author: Oscar Esteban
% @Date:   2015-02-20 15:40:09
% @Last Modified by:   Oscar Esteban
% @Last Modified time: 2015-03-18 15:08:42

@@

% *** MISC UTILITY PACKAGES ***
%\usepackage{fancyhdr}
\usepackage{multicol}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs} % to use \toprule
\usepackage{makeidx}
%\usepackage[scale=2.0]{ccicons}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[top=2cm,bottom=1.8cm,left=2.1cm,right=1.7cm,marginparwidth=1.2in]{geometry}
\usepackage{changepage}
\usepackage{epigraph}
\usepackage[toc,nomain,acronym,shortcuts,translate=false]{glossaries}

\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{cancel}
%\usepackage{algorithmic}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\const}{Const.}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\providecommand{\mdist}[2]{ \mathcal{D}_{#2}^2(\mathbf{#1}) }
\let\oldhat\hat
\renewcommand{\vec}[1]{\mathbf{#1}}
\providecommand{\nvec}[1]{\hat{\mathbf{#1}}}

\renewcommand\thesection{S\arabic{section}}
\renewcommand\thesubsection{S\arabic{section}.\arabic{subsection}}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand{\theequation}{SM\arabic{equation}}

@@

\usepackage[hyphens]{url}
\usepackage[hyphenbreaks]{breakurl} %fixes boxes spanning through pages
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue,linkcolor=blue]{hyperref}

\input{01-acronyms}

@@

\maketitle
\section{Extensions to the mathematical formulation of the methods}

\paragraph{Discretization of the shape gradients} The energy functional to be optimized
  in \emph{regseg} is presented in Eq. 7:
  \begin{equation}
  E(R \mid U) = C + \underbracket{\underset{l}{\sum} \int_{\Omega_l}
  \mdist{f'}{l} \,d\vec{r}}_{\text{Data term}}
  + \underbracket{\underset{\Omega}{\int} \left[ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} \right] \,d\vec{r}}_{\text{Regularization term}}.
  \label{eq:energy}
  \end{equation}

To search for the minimum of $E(R \mid U)$ w.r.t. the coefficients $\vec{u}_k$, we use a gradient descent strategy.
In Eq. 11 we introduced the derivative of \eqref{eq:energy}:
  \begin{equation}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} =
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r}
  + \int_{\Omega} [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{equation}

Let us compute the derivative of the data term first (Eq. 12):
\begin{equation}
  \frac{\partial E_{data}(\vec{u})}{\partial \vec{u}_k} =
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \Big\} =
  \underset{l}{\sum} \, \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \int_{\Omega_l} \mdist{f'}{l} \Big\},
  \label{eq:data_derivative}
\end{equation}
where we can introduce the shape-gradients theory from Appendix 1.
The equation A.1 is discretized as follows.
First, the surface between limiting regions $\{l, m\}$, $\Gamma_{l,m}$ is explicitly represented by
  a discrete set of vertices $\vec{v}_i$, with $i \in \{0, \ldots, N_p -1 \}$.
Complementarily, the inwards normal of the surface $N_{\Gamma_{l,m}}$ is represented by the discrete
  set of normals $\hat{\vec{n}}_i$ at each vertex of the mesh.
The resulting summation is, therefore, discrete and the integral operator is replaced by the sum.
  \begin{align}
  \frac{\partial}{\partial t} \int_\Omega F(\vec{r}) d\vec{r} &=
  \underbracket{\cancel{\int_\Omega \frac{\partial}{\partial t}F(\vec{r}) d\vec{r} }}_{\text{Functional's evolution}}
  - \underbracket{\int_{\Gamma_{l,m}} F(\vec{r}) \left\langle \frac{\partial \Gamma_{l,m}}{\partial t},
  N_{\Gamma_{l,m}}\right\rangle d\vec{r}}_{\text{Shape's evolution}} \notag \\
  & = - \underset{p}{\sum} \frac{1}{A_p} \underset{i}{\sum} \, a_i \, F(\vec{v}_i) \left\langle \underbracket{\frac{\partial \vec{v}_i}{\partial t}}_{\text{speed of }\vec{v}_i},
  \hat{\vec{n}}_{i}\right\rangle.
  \label{eq:shape_gradient_orig}
  \end{align}
where $a_i$ is the area corresponding to vertex $\vec{v}_i$, and $A_p = \sum_i a_i$ is the total area of surface $p$.
In the following, we will refer as $w_{p,i} = a_i / A_p $ to the area contribution of $\vec{v}_i$ to the
  total area of the surface it belongs to.
For simplicity, the sum over $p$ can be also removed, as the vertices belong to only one of the total $P$ contours

Then, the speed of $\vec{v}_i$ is discretized using the artificial timestep parameter $\delta$, as the displacement
  $\vec{d}_i = \vec{v}_i(\delta = t+1) - \vec{v}_i(\delta = t)$:
  \begin{equation}
  \frac{\partial}{\partial t} \int_\Omega F(\vec{r}) \, d\vec{r} =
  - \underset{i}{\sum} w_{p,i} F(\vec{v}_i) \vec{d}_i \cdot \hat{\vec{n}}_i.
  \label{eq:shape_gradient_disc1}
  \end{equation}

Since the energy functional is defined over competing regions, the displacement of $\vec{v}_i$ will cause
  an energy exchange between the limiting regions, and therefore $F(\vec{r})$ must be splitted in
  two terms, $F_{in}(\vec{r})$ corresponding to the interior region and $F_{out}(\vec{r})$ to the exterior:
  \begin{equation}
  \frac{\partial}{\partial t} \int_\Omega F(\vec{r}) \, d\vec{r} =
  - \underset{i}{\sum} w_{p,i} \, \vec{d}_i \cdot
  \underbracket{\Big[ \big(F_{out}(\vec{v}_i) - F_{in}(\vec{v}_i) \big) \hat{\vec{n}}_i \Big]}_{\bar{s}_i \text{ in Figure 1}}.
  \label{eq:shape_gradient_disc2}
  \end{equation}

Now we can identify $t$ with the set of time-dependent coefficients, $\{\vec{u}_k\}$, and the target function $F$ with $\mathcal{D}_l^2$.
To introduce the speed of the vertices from equation 13, we replace $\vec{d}_i$:
\begin{equation}
  \frac{\partial E_{data}(\vec{u})}{\partial \vec{u}_k} =
  - \underset{i}{\sum} w_{p,i} \, \psi_k(\vec{v}_i) \hat{\vec{e}} \cdot \bar{s}_i =
  - \underset{i}{\sum} w_{p,i} \, \psi_k(\vec{v}_i) \bar{s}_i = - \underset{i}{\sum} \vec{g}_{i,k}.
  \label{eq:shape_gradient_final}
\end{equation}

Inserting \eqref{eq:shape_gradient_final} in \eqref{eq:gradient_descent} we reach equation 15,
  and renaming $\vec{g}_k = - \underset{i}{\sum} \vec{g}_{i,k}$:
  \begin{equation}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} =
  \vec{g}_k + \boldsymbol{\alpha} \vec{u}_k
  -\boldsymbol{\beta} \Delta \vec{u}_k.
  \label{eq:gradient_final}
  \end{equation}

\paragraph{Obtaining the update equation} Once the shape gradients are obtained \eqref{eq:gradient_final},
  we apply the Fourier Transform to solve the \gls*{pde}:
\begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} \equiv \frac{\vec{u}_k^{t+1} - \vec{u}_k^t}{\delta} \implies
  \vec{u}_k^{t+1} &= \delta \Big( \vec{g}_k + \boldsymbol{\alpha} \vec{u}_k
  - \boldsymbol{\beta} \Delta \vec{u}_k \Big) + \vec{u}_k^t, \notag \\
  \vec{u}_k^{t+1} &= \delta \vec{g}_k +  \boldsymbol{\alpha} \delta \vec{u}_k
  - \boldsymbol{\beta} \delta \Delta \vec{u}_k + \vec{u}_k^t
  \label{eq:discrete_derivative}
\end{align}

@@

\paragraph{Efficient interpolation using sparse matrices}
During the registration process, every iteration requires computing the product of all the gradients
  $\vec{g}_k$ associated to the control point $k$, and computed at the current position
  $\vec{v}_i$ by the corresponding weights $\psi_{ik} = \psi_k(\mathbf{v}_i)$ of interpolating functions
  \eqref{eq:shape_gradient_final}.
In order to optimize multiplications and summations, all the $\psi_{ik}$ are collected in a
  matrix $\boldsymbol{\Psi} = (\psi_{ik})$.
Given the limited support of the basis function $\psi$, $\boldsymbol{\Psi}$ will hold the property
  of being sparse, as only few $\psi_{ik} > 0$ in the surroundings of $\vec{v}_i$.
Then, the gradients $g_k$ are easily computed using the matrix product:
\begin{equation}
  \big(\vec{g}_k \big) = \boldsymbol{\Psi} \cdot \big( w_i \, \bar{s}_i \big)^T
  \label{eq:sparse_matrix}
\end{equation}

As these weights can be computed once in the beginning of the process and they do not change along
  it, $\boldsymbol{\Psi}$ can be pre-cached.

@@

\subsection{Convergence evidencing}\label{sec:convergence_evidence}

In order to track the evolution of the registration process, several internal variables
  are saved in the JSON log-file.
Using the JSON log-file as input for the \emph{nipype Interface}
  \codeinline{ACWEReport}, it is straightforward to obtain
  a visual assessment document presenting the convergence.
\begin{pythoncode}
from pyacwereg.interfaces.acwereg import ACWEReport
report = ACWEReport()
report.inputs.in_log = `myprefix.log'
ifresult = report.run()
\end{pythoncode}

Online checking is also possible as the algorithm writes to the standard output as well.
A sample report is found in \autoref{fig:convreport}.

\begin{figure*}[b]
  \includegraphics[width=\textwidth]{figures/Suppl-figure02.pdf}
  \caption{The evolution of the registration and segmentation process can be
    checked using the \emph{Convergence report},
    easily generated using the appropriate \emph{nipype Interface}.
  The report comprehends several plots tracking the evolution of the algorithm and several
    features to help researchers tune up the algorithm in their application.}%
    \label{fig:convreport}
\end{figure*}

@@

\newpage
\section{Extended results on real data}

\immediate\write18{./visual_report.sh}
\input{suppl-realdata.tex}
